---
title: "Hospital cost analysis"
output: pdf_document
---


DESCRIPTION

Background and Objective:

A nationwide survey of hospital costs conducted by the US Agency for Healthcare consists of hospital records of inpatient samples.
The given data is restricted to the city of Wisconsin and relates to patients in the age group 0-17 years.
The agency wants to analyze the data to research on healthcare costs and their utilization.

Domain: Healthcare

Dataset Description:


Attribute	     Description
Age 	         Age of the patient discharged
Female 	       A binary variable that indicates if the patient is female
Los	           Length of stay in days
Race 	         Race of the patient (specified numerically)
Totchg	       Hospital discharge costs
Aprdrg         All Patient Refined Diagnosis Related Groups

```{r}
# Loading all the packages required.

library(readxl)
library(dplyr)
library(ggplot2)

# Importing the dataset

setwd("C:/DataScience/Rprojects/Hospitalanalysis")
getwd()
hospitaldata<-read_xlsx("hospitalcosts.xlsx")
View(hospitaldata)


# Q1)To record the patient statistics, the agency wants to find the age 
# category of people who frequent the hospital and has the maximum expenditure.


agecount<-hospitaldata %>% count(AGE)
View(agecount)

ageg<-ggplot(hospitaldata)+ geom_bar(aes(x = AGE), fill="red")
ageg


# we see that most frequent age of discharge is 0 with a count of 307
# (61.4% in the given data set) . So we can say babies under age 1
# that is babies of age 0 are more frequent to the hospital.

ageexp<-hospitaldata %>% count(AGE, wt= TOTCHG)
View(ageexp)


# Maximum expenditure is of Patients of age 0, this might be because 61.4% of
# the dataset are age 0 patient, so we can take average expenditure of each age
# to find the age with maximum average expenditure.

summ<-summarise(group_by(hospitaldata,AGE),mean(TOTCHG))
View(summ)

plot(summ,col="red",bg="red")

#maximum average expenditure is from the patients of age 9.

# Answer 1

# ->From the data set we conclude Age 0 is more frequent with
# 61.4% of all the patients. 
# ->Patients with Maximum expenditure is age 0
# also to note maximum average expenditure belongs to patient of age 9.


```


```{r}
# Q2)In order of severity of the diagnosis and treatments and to find out
# the expensive treatments, the agency wants to find the diagnosis-related group
# that has maximum hospitalization and expenditure.


drgcount<-hospitaldata %>% count(APRDRG)
View(drgcount)

# The DRG which has maximum hospitalization is
# 640 (Hemorrhage in early pregnancy) with 267 patients.

drgavgexp<-summarise(group_by(hospitaldata,APRDRG),mean(TOTCHG))
View(drgavgexp)
drgsumexp<-summarise(group_by(hospitaldata,APRDRG),sum(TOTCHG))
View(drgsumexp)

# so we see the maximum average expenditure is DRG 911 with an average of 48388
# and the maximum sum of expenditure is for the DRG 640,
# this is  because DRG 640 is 53.4% of all the DRGs in the given dataset.


# Answer 2

# From the data set we conclude that the DRG which has maximum hospitalization
# is 640, 53.4% of the total patients. DRG 640 contributes to the
# maximum expenditure and we can also note the
# maximum average expenditure is for DRG 911.
```



```{r}
# Q3)To make sure that there is no malpractice, the agency needs to analyze
# if the race of the patient is related to the hospitalization costs.



racec<-hospitaldata%>% count(RACE)
View(racec)
raceexp<-summarise(group_by(hospitaldata,RACE),mean(TOTCHG))
View(raceexp)

# In the given dataset the most frequent patients belong to race 1 and
# the maximum average expenditure is of race 2. With this data we can not say
# Race of a patient affects the cost as other factors might have
# effected the cost.




racexplm<-lm(data=hospitaldata, formula = TOTCHG~RACE)
summary(racexplm)

# By doing the linear regression we can say that TOTCHG (Hospitalization cost)
# in not dependent on the race of the patient for the given data set.

#Answer 3

# -> It seemed like the cost was affected by race of the patient. But by doing
# the linear regression we see that the p value is greater than 0.05 hence race
# of a patient does not affect the cost.
```


```{r}
# Q4)To properly utilize the costs, the agency has to analyze the
# severity of the hospital costs by age and gender for the proper allocation
# of resources.



selection<-select(hospitaldata,AGE,FEMALE,TOTCHG)
View(selection)
t<-summarise(group_by(selection,AGE,FEMALE),mean(TOTCHG))
View(t)


facfemale<-factor(selection$FEMALE)
ggplot(hospitaldata, aes(fill=facfemale, y=TOTCHG, x=AGE)) + 
    geom_bar(position="dodge", stat="identity")

# By looking at the grouped bar chart we can say for the age 0,5,12,13,17
# females have higher expenditure than males of those age.

hostag<-lm(hospitaldata, formula= TOTCHG~ AGE+ FEMALE)
summary(hostag)
hostag$coefficients

#Answer 4

# ->From the bar chart we can say that Females of ages 0,5,12,13,17 have higher
# expenditure than males of those age.


# ->From the linear regression we can conclude that AGE is affecting the
# cost more than GENDER.We get the equation
# y = 2718.62816 + 86.28231*AGE - 748.18748*FEMALE.



```


```{r}
# Q5) Since the length of stay is the crucial factor for inpatients,
# the agency wants to find if the length of stay can be predicted
# from age, gender, and race.

length<-lm(hospitaldata, formula = LOS~ AGE + FEMALE + RACE)
summary(length)


# Answer 5

# -> We can see that the p value is high for all three variables which implies
# the linear relationship  is not established between dependent variable
# LOS and Independent variable (AGE,GENDER,RACE). Hence LOS is not predictable
# using AGE, FEMALE, RACE from the given data.
 

```


```{r}

#  Q6) To perform a complete analysis, the agency wants to find the variable
# that mainly affects hospital costs.

cost<-lm(hospitaldata, formula = TOTCHG ~ APRDRG +AGE +FEMALE +LOS +RACE)
summary(cost)

# Answer 6

# -> by looking at the p value for each of the category we can remove
# gender and race as their p value is high, meaning they do not affect Cost
# in the given data set. 
# -> Hence we can consider Length Of Stay ,AGE and APRDRG which impacts
# the cost.


```

